\textbf{Real-time guitar transcription using\\Fourier transform based methods}\\A pitch estimation framework and overtone sieve algorithm

### 2021-04-14 paper.tex
future work en abstract

\section*{Abstract}
% In the field of automatic music transcription (AMT), data driven methods such as artificial intelligence and machine learning are becoming more popular by the year. This is especially true for multi-pitch estimation (also known as multiple $f_0$ estimation). Even though the results are getting better every year, little understanding of the problem is gained. The few papers which use mathematical models to solve this problem, do not publish their code which makes it difficult to build upon.

@@

Pitch estimation can be divided into a few subproblems. If a paper focusses on one of these subproblems, it also has to solve the other subproblems in order to test the system. Most papers do not publish their source code, which makes building upon other's work difficult. It also limits the research of the interaction between different solutions for the different subproblems. Because of this, we provide a framework in which can be built upon, as the implementation of a subproblem can easily be interchanged with another. The code can be found at "github.com/lucmans/dechord".

In short, the goal of this paper is developing a program which can translate a guitar signal to a set of observed notes in real-time. We also provide an algorithm which can estimate a set of played notes using the set of observed notes. The code should be publically available and easily extensible so others can optimize a subproblem without having to solve the other subproblems.

@@

A big problem in polyphonic pitch estimation is the occurrence of overtones~\cite{oud}. As these overtones also have a periodic nature in the frequency domain, they can be detected using another Fourier transform~\cite{double}. This does however have a big impact on the latency, as performing a Fourier transform is computationally expensive. Also, notes which are at octave intervals have the same periodicity~\cite{octave}, which makes them difficult to distinguish.

@@

\subsection{Latency}  \label{sub:algspeed}
The processing time of our system is very important, as it will directly impact the latency. We will only measure the time between receiving a frame from the audio driver until the answer is computed, as the time spent waiting on enough samples is fully determined by the number of samples in a frame and the sampling rate. The total processing time and the processing time of the individual steps and the total system can be seen in Figure~\ref{fig:snel}. Here, we only measure the latency on frames which contain peaks.
\begin{figure}[t]%\vspace{-2mm}
    \centering
    \includegraphics[width=\linewidth]{1.png}\vspace{+6mm} %0.85
    \includegraphics[width=\linewidth]{2.png} %0.85
    \caption{Processing time of the different algorithms}
    \label{fig:snel}
\end{figure}%\vspace{-4mm}
The total latency is never higher than 360 microseconds. This is well within out real-time constraint if we disregard the latency to fill a frame. Surprisingly, the peak picking algorithm is more compute intensive than the transform. Most of the time, the latency is on the lower end of the distribution. The outliers are mostly a result of transients. During transients, there are a lot more peaks to process, which slows down the different algorithms. This is especially the case for polyphonic transcription, as we have to take all the peaks into account. In the polyphonic dataset, transients contain more peaks as multiple strings are strummed. This results in a second lobe in the mid segment of the latency distribution.

@@

All code related to this project can be found at "github.com/lucmans/dechord".

\section{Future work}  \label{sec:future}
How to make results more consistent (onset detection, transient filtering). How to make the project more real-time (threading rolling transforms). Also deliberately detuning strings for better results. Amplification->more overtones detected but also more noise. Remove spikes in latency by saving FFTW wisdom.

This algorithm has some problems in its current form. Solving these is outside of the scope of this paper and will be the topic of the follow-up thesis, but we will address the problems here. The first big problem is transients. They will appear as many random peaks in the frequency domain. Many of these peaks will be outside of the fundamental frequency range of a guitar, but not all of them, so discarding all high peaks does not solve the problem. Usually, during transients, there any significantly more peaks and even multiple peaks within one note. If this can pattern can consistently be detected, samples can be discarded until the transient is filtered. Another problem is the resolution in the lower frequencies.

\subsection{Multi-threading}  \label{sub:futlaten}
asdf

>>>>

\section*{Abstract}
In this paper, we explore real-time transcription of a electric guitar signal. Data driven methods are not fast enough for real-time use, so we focus on Fourier transform based methods. We will provide a basis for monophonic transcription and a proof-of-concept algorithm for polyphonic transcription. The goal of this paper is to give an overview of the problems when trying to solve this problem.
% In the field of automatic music transcription (AMT), data driven methods such as artificial intelligence and machine learning are becoming more popular by the year. This is especially true for multi-pitch estimation (also known as multiple $f_0$ estimation). Even though the results are getting better every year, little understanding of the problem is gained. The few papers which use mathematical models to solve this problem, do not publish their code which makes it difficult to build upon.

@@

Pitch estimation can be divided into a few subproblems. If a paper focusses on one of these subproblems, it also has to solve the other subproblems in order to test the system. Most papers do not publish their source code, which makes building upon other's work difficult. It also limits the research of the interaction between different solutions for the different subproblems. Because of this, we provide a framework in which can be built upon, as the implementation of a subproblem can easily be interchanged with another. The code can be found at "\url{github.com/lucmans/dechord}".

The goal of this paper is to give an overview of the problems when trying to solve this problem. This is done by developing a program which can translate a guitar signal to a set of observed notes in real-time. We also provide an algorithm which can estimate a set of played notes using the set of observed notes. The code should be publically available and easily extensible so others can optimize a subproblem without having to solve the other subproblems.

@@

A big problem in polyphonic pitch estimation is the occurrence of overtones~\cite{oud}. As these overtones also have a periodic nature in the frequency domain, they can be detected using another Fourier transform~\cite{double}. This does however have a big impact on the latency, as performing a Fourier transform is computationally expensive. Also, notes which are at octave intervals have the same periodicity~\cite{octave}, which makes them difficult to distinguish. This is called the octave problem.

@@

\subsection{Latency}  \label{sub:algspeed}
The processing time of our system is very important, as it will directly impact the latency. We will only measure the time between receiving a frame from the audio driver until the answer is computed, as the time spent waiting on enough samples is fully determined by the number of samples in a frame and the sampling rate. The total processing time and the processing time of the individual steps and the total system can be seen in Figure~\ref{fig:snel}. Here, we only measure the latency on frames which contain peaks.
\begin{figure}[t]%\vspace{-2mm}
    \centering
    \includegraphics[width=\linewidth]{1.png}\vspace{+6mm} %0.85
    \includegraphics[width=\linewidth]{2.png} %0.85
    \caption{Processing time of the different algorithms}
    \label{fig:snel}
\end{figure}%\vspace{-4mm}
The total latency is never higher than 360 microseconds. In other words, we are at least 240 times real-time. This is well within out real-time constraint if we disregard the latency to fill a frame. Surprisingly, the peak picking algorithm is more compute intensive than the transform. Most of the time, the latency is on the lower end of the distribution. The outliers are mostly a result of transients. During transients, there are a lot more peaks to process, which slows down the different algorithms. This is especially the case for polyphonic transcription, as we have to take all the peaks into account. In the polyphonic dataset, transients contain more peaks as multiple strings are strummed. This results in a second lobe in the mid segment of the latency distribution.

@@

Our peak picking algorithm works fine for monophonic transcription, however, it does not work too well for polyphonic transcription. For instance, peaks in close distance to each other are easily missed. It is also compute intensive when compared to the other components and has the largest outliers. Even though the outlier are relatively very severe, they are all within 200 microseconds and might not be noticeable.

All code related to this project can be found at "\url{github.com/lucmans/dechord}".

\section{Future work}  \label{sec:future}
Our program serves as a good starting point for further research. However, there are still some problems with our current implementation. Solving these is outside of the scope of this paper and will be the topic of the follow-up thesis, but we will address the problems here. The first big problem is transients. They will appear as many random peaks in the frequency domain. Many of these peaks will be outside of the fundamental frequency range of a guitar, but not all of them, so discarding all high peaks does not solve the problem. Usually, during transients, there any significantly more peaks and even multiple peaks within one note. If this can pattern can consistently be detected, samples can be discarded until the transient is filtered.

The Fourier transform requires too many samples in a frame to remain within the real-time constraint while still providing the necessary frequency resolution in the fundamental frequencies. It might be possible to alleviate this problem with other transforms, such as the constant-Q transform or the wavelet transform, but those transforms may have to solve other problems. As mentioned before, the low resolution does not have to be a problem, as less resolution is needed when discerning higher pitched peaks. Currently, we do not detect enough overtones, but some testing showed us we can detect many overtones if we amplify the signal. This has to be done carefully, as artifacts may be introduced in the signal. Optimally amplifying of a recording is trivial, however, it is trickier in real-time. Using amplified signals would require recalibration of the peak picking parameters or even require a new algorithm in general.

The current implementation has no onset detection. We stated before that onset detection may help to filter out transients. If we can detect the transient in the time domain, we can discard all samples containing the transient. This may however be very difficult to do in real-time. If we can detect that there was a transient in a transformed frame, we are too late to act on it due to the large frame times. We can circumvent this by transforming partially overlapping frames. This could be done in different threads to limit the extra computational pressure which could impact our latency. However, as we are about 240 times real-time, we could likely perform it single threaded. Now, if we have a bad frame, we have to wait a full frame (83 ms) for the next estimation. However, if we use $n$ partially overlapping frames, we can try another estimation in $83/n$ ms. This will increase the responsiveness of our system greatly.

It is virtually impossible to know on what string a certain note was played. Furthermore, when playing an octave, the overtones of the lower pitched note align perfectly with the higher note and its overtones. If the strings were deliberately detuned to orthogonal frequencies, we can easily distinguish between the played strings. This does however create more peaks in the frequency domain. This tradeoff is worth further research.
